# Chapter 6: Analysis of longitudinal data

```{r}
date()
```

## Data Analysis

```{r}
# Read in data
BPRSL <- read.table("data/BPRSL_data.csv",
                    sep =",", header = T)

RATSL <- read.table("data/RATSL_data.csv",
                    sep =",", header = T)

# 2. Categorical variables to factors

# Factor treatment & subject
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

# Factor ID and Group
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

```

Here we have two data sets:

-   Data on (medical) treatments for human subjects

-   Data on rats' diets

### Part I: Graphical dissection of rats' diets

#### Rats' diets

```{r}
str(RATSL)
summary(RATSL)

dim(RATSL)

```

```{r}
# Glimpse the data
glimpse(RATSL)

```

There are 16 subjects in 3 groups. The *Time* variable indexes the first day of the week (initial status is day 1).

```{r}
library(ggplot2)

# Draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

For the first group, the weight is low to begin with and keeps constant. For the second and third groups, the initial weights are higher and keep rising with time. The second group seems to have one outlier. The other groups may also have them.

Standardize variables:

```{r}
RATSL_std <- RATSL %>%
  group_by(Time) %>%
  mutate(stdwt = Weight) %>%
  ungroup()

glimpse(RATSL_std)
```

```{r}

# Draw the plot
ggplot(RATSL_std, aes(x = Time, y = stdwt, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL_std$Weight), max(RATSL_std$Weight)))

```

Does anything change?

#### Error bars and outliers

```{r}
# Summary data with mean and standard error of weight by group and time
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = Weight, se = Weight ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL)

# Glimpse the data
glimpse(RATSS)

# Plot the mean profiles
library(ggplot2)
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  #theme(legend.position = c(0.8,0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")

```

Unfortunaltely, this plot doesn't look very useful.

Are there any outliers?

```{r}
RATSL10S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

glimpse(RATSL10S)

ggplot(RATSL10S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 1-10")

```

Each group seems to have one outlier. Let's remove them.

```{r}
RATSL10S1 <- RATSL10S %>%
  filter(Group == 1, mean > 250)
glimpse(RATSL10S %>% filter(Group == 1))
glimpse(RATSL10S1)
RATSL10S2 <- RATSL10S %>%
  filter(Group == 2, mean < 550)
glimpse(RATSL10S %>% filter(Group == 2))
glimpse(RATSL10S2)
RATSL10S3 <- RATSL10S %>%
  filter(Group == 3, mean > 500)
glimpse(RATSL10S %>% filter(Group == 3))
glimpse(RATSL10S3)
RATSL10SF = rbind(RATSL10S1, RATSL10S2, RATSL10S3)
glimpse(RATSL10SF)
```

```{r}
ggplot(RATSL10S, aes(x = Group)) +
  geom_bar()
```

```{r}
ggplot(RATSL10SF, aes(x = Group)) +
  geom_bar()
```

The data, replotted:

```{r}
ggplot(RATSL10SF, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 1-10")
```

#### Testing distribution

T-test: check whether there are differences between the distributions of two groups. Since there are three groups in this case, we must do three pair-wise comparisons.

The tests confirm that the distributions are different (based on p-values).

```{r}
t.test(mean ~ Group, data = RATSL10SF %>% filter(Group != 1), var.equal = TRUE)
t.test(mean ~ Group, data = RATSL10SF %>% filter(Group != 2), var.equal = TRUE)
t.test(mean ~ Group, data = RATSL10SF %>% filter(Group != 3), var.equal = TRUE)

```

ANOVA analyses differences in variance. The baseline (initial weight) is highly significant for the result.

```{r}
library(dplyr)
library(tidyr)
# Add the baseline from the original data as a new variable to the summary data
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')
RATSL10S2 <- RATSL10S %>%
  mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSL10S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)

```

### Part II: Human treatments

#### Medical treatments

```{r}

str(BPRSL)
summary(BPRSL)

dim(BPRSL)

```

There are 20 subjects in two groups (*treatment*). The experiment lasted for 8 weeks (+ the additional *week 0* as the starting condition).

```{r}
str(BPRSL)
```

```{r}
# Draw the plot
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

```

The plots show generally effective treatment for both groups.

#### Dumb Linear Model

Fitting a linear model to time-series data isn't generally a good idea, since the observations obviously are not independent (for a specific subject). We'll try anyway.

```{r}
# create a regression model BPRS_reg
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)

# print out a summary of the model
summary(BPRS_reg)
```

#### Linear Mixed Effect Model

An individual's response may depend on a variety of things, some of them unobserved. These are included in the model as random variables.

We'll begin with fitting a *random intercept model* with *week* and *treatment* as explanatory variables, with randomness added for each subject.

```{r}
library(lme4)

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)
```

Note that the standard error is higher for the intercept and lower for week and treatment. The lower error for the time-dependent variable is expected, as values are expected to vary less at a specific point in time across individuals and groups. Lower error for treatment is interesting, which would suggest that there are clear differences in the effectiveness of treatments, when taken random effects into account.

#### Random Intercept and Random Slope model

Allows a random effect to have slopes in the linear regression.

```{r}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)
```

NB: Not sure how to interpret this.

```{r}
BPRS_ref2 <- lmer(bprs ~ week + treatment + (week | treatment), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

```

```{r}
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)

```

```{r}
BPRSL$fitted <- Fitted

str(BPRSL)
str(Fitted)

ggplot(BPRSL, aes(x = week, y = Fitted, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

```

Both treatments appear to be successful.
